---
title: "Lesson 0: Primer in ecological thinking"
author: "Alyssa Willson"
date: "13/5/2020"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(viridis)
library(deSolve)
```

## Objectives

1. Become familiar with some basic ecological questions

2. Consider how to answer the questions identified

3. Define the quantitative information required for each example

## Today we will consider:

1. Population ecology: dynamics of one species through time

2. Community ecology: dynamics of multiple species through time

3. Disease ecology: progression of an epidemic through time

## Population ecology

* The question: How does the size of a population change over time?

* The context: the number of seals in Washington

* The method: ??

    + $\textbf{How do we think the number of seals will change}$
      $\textbf{over time?}$
    
    + $\textbf{Draw the curve. We will revisit how to represent}$
      $\textbf{this mathematically later.}$
      
## Population ecology: problem visualization

```{r, echo = F}
r = 0.2
K = 7500
x0 = 1600
nstep = 50
x = rep(0, nstep)
x[1] = x0

for(i in 2:nstep){
  x[i] = x[i - 1] + r * x[i - 1] * (1 - x[i - 1] / K)
}

ggplot() +
  geom_line(aes(x = 1:nstep, y = x), size = 1.25) +
  xlab('Time') + ylab('Number of seals')
```

## Population ecology: analysis

```{r,  out.width= "50%", out.extra='style="float:right; padding:10px"'}
ggplot() +
  geom_line(aes(x = 1:nstep, y = x), size = 1.25) +
  xlab('Time') + ylab('Number of seals')
```

</div>

* Under what circumstances does the population increase?

* Stay the same?

* Could the population ever decrease?

## Community ecology

* The question: How do predator and prey populations affect each other over time?

* The context: classic example from Isle Royale, Michigan

    + Predator: wolves
  
    + Prey: moose
  
* The method: ??

    + $\textbf{Do we expect the size of one species' population to}$
      $\textbf{affect the other?  If so, how?}$
    
    + $\textbf{Draw a curve for each of the species on the same}$
      $\textbf{graph}$
      
## Community ecology: problem visualization

```{r}
Pars <- c(alpha = 1.5, beta = .1, gamma = .5, delta = .1)
State <- c(x = 10, y = 10)

LotVmod <- function (Time, State, Pars) {
  with(as.list(c(State, Pars)), {
    dx = x*(alpha - beta*y)
    dy = -y*(gamma - delta*x)
    return(list(c(dx, dy)))
  })
}

out = ode(func = LotVmod, y = State, parms = Pars, times = seq(1, 50, 0.1))

ggplot() +
  geom_line(aes(x = out[,1], y = out[,2], color = 'Prey'), size = 1.25) +
  geom_line(aes(x = out[,1], y = out[,3], color = 'Predator'), size = 1.25) +
  xlab('Time') + ylab('Population size') +
  theme(legend.title = element_blank())
```

## Community ecology: analysis

```{r,  out.width= "50%", out.extra='style="float:right; padding:10px"'}
ggplot() +
  geom_line(aes(x = out[,1], y = out[,2], color = 'Prey'), size = 1.25) +
  geom_line(aes(x = out[,1], y = out[,3], color = 'Predator'), size = 1.25) +
  xlab('Time') + ylab('Population size') +
  theme(legend.title = element_blank())
```

</div>

* When would you expect the prey population to increase? to decrease?

* When would the predator population increase and decrease?

* What is the general relationship between these two populations?

## Disease ecology

* The question: how does the number of people susceptible to the disease, infected with the disease, and recovered from the disease change over time?

* The context: seasonal influenza in a community over the course of one flu season

* The method: ??

    + $\textbf{How will each pool (susceptible, infected, recovered)}$
      $\textbf{change over time?}$
    
    + $\textbf{Draw one possible scenario for each pool.}$
    
## Disease ecology: problem visualization

```{r}
beta = 1
gamma = 0.5
State = c(S = 10, I = 1, R = 0)
Pars = c(beta, gamma)

SIR = function(time, State, Pars){
  with(as.list(c(State, Pars)),{
    N = S + I + R
    dS = -beta * (S * I) / N
    dI = (beta * (S * I) / N) - gamma * I
    dR = gamma * I
    return(list(c(dS, dI, dR)))
  })
}

out = ode(func = SIR, y = State, parms = Pars, times = seq(1, 15, 0.1))

colors = viridis(3)
ggplot() +
  geom_line(aes(x = out[,1], y = out[,2], color = 'S'), size = 1.25) +
  geom_line(aes(x = out[,1], y = out[,3], color = 'I'), size = 1.25) +
  geom_line(aes(x = out[,1], y = out[,4], color = 'R'), size = 1.25) +
  xlab('Time') + ylab('Fraction of population') +
  scale_color_manual(values = c('S' = colors[1], 'I' = colors[2], 'R' = colors[3]), name = '', breaks = c('S', 'I', 'R'))
```

## Disease ecology: analysis

```{r,  out.width= "50%", out.extra='style="float:right; padding:10px"'}
ggplot() +
  geom_line(aes(x = out[,1], y = out[,2], color = 'S'), size = 1.25) +
  geom_line(aes(x = out[,1], y = out[,3], color = 'I'), size = 1.25) +
  geom_line(aes(x = out[,1], y = out[,4], color = 'R'), size = 1.25) +
  xlab('Time') + ylab('Fraction of population') +
  scale_color_manual(values = c('S' = colors[1], 'I' = colors[2], 'R' = colors[3]), name = '', breaks = c('S', 'I', 'R'))
```

</div>

* Describe how each curve changes over time.

* Can the "S" pool ever increase?  Can the "R" pool ever decrease?

* What does the "I" pool represent?

